\section{Architecture}
\label{sec:architecture}

The \texttt{mqtt5} library maps MQTT protocol operations onto QUIC streams using configurable \emph{stream mapping strategies}. The architecture decouples the MQTT session layer from the transport layer, allowing the same MQTT implementation to operate over TCP, TLS~1.3, WebSocket, or QUIC with different stream configurations.

\subsection{Stream Mapping Strategies}

We define three strategies for mapping MQTT control and data packets onto QUIC streams, each offering a different point in the isolation--overhead tradeoff space.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    stream/.style={draw, rounded corners=2pt, minimum width=1.6cm, minimum height=0.5cm, font=\footnotesize},
    connbox/.style={draw, dashed, rounded corners=4pt, inner sep=6pt},
    label/.style={font=\footnotesize\bfseries},
    >=Stealth
]

\begin{scope}[shift={(0,0)}]
\node[label] at (0.8, 2.0) {(a) Control-only};
\node[connbox, minimum width=2.2cm, minimum height=2.8cm] at (0.8, 0.2) {};
\node[stream, fill=blue!15] at (0.8, 1.2) {Stream 0};
\node[font=\tiny, align=center] at (0.8, 0.6) {CONNECT, PUBLISH,\\SUBSCRIBE, \ldots};
\node[font=\tiny, text=gray] at (0.8, -0.4) {All on one stream};
\end{scope}

\begin{scope}[shift={(3.2,0)}]
\node[label] at (0.8, 2.0) {(b) Per-topic};
\node[connbox, minimum width=2.2cm, minimum height=2.8cm] at (0.8, 0.2) {};
\node[stream, fill=blue!15] at (0.8, 1.2) {Stream 0};
\node[stream, fill=green!15] at (0.8, 0.5) {Stream 1};
\node[stream, fill=orange!15] at (0.8, -0.2) {Stream 2};
\node[font=\tiny] at (0.8, 1.55) {\textit{control}};
\node[font=\tiny] at (0.8, 0.85) {\textit{topic/a}};
\node[font=\tiny] at (0.8, 0.15) {\textit{topic/b}};
\end{scope}

\begin{scope}[shift={(6.4,0)}]
\node[label] at (0.8, 2.0) {(c) Per-publish};
\node[connbox, minimum width=2.2cm, minimum height=2.8cm] at (0.8, 0.2) {};
\node[stream, fill=blue!15] at (0.8, 1.2) {Stream 0};
\node[stream, fill=yellow!20] at (0.8, 0.5) {Stream $n$};
\node[stream, fill=yellow!20] at (0.8, -0.2) {Stream $n$+1};
\node[font=\tiny] at (0.8, 1.55) {\textit{control}};
\node[font=\tiny] at (0.8, 0.85) {\textit{msg 1}};
\node[font=\tiny] at (0.8, 0.15) {\textit{msg 2}};
\end{scope}

\end{tikzpicture}
\caption{The three stream mapping strategies. (a)~Control-only multiplexes everything on stream~0. (b)~Per-topic assigns a dedicated stream per topic, providing topic-level isolation. (c)~Per-publish creates a new stream for each PUBLISH, providing message-level isolation.}
\label{fig:strategies}
\end{figure}

\textbf{Control-only} (Fig.~\ref{fig:strategies}a) uses a single bidirectional QUIC stream for all MQTT traffic. This is functionally equivalent to running MQTT over a single TCP connection and provides no HOL blocking mitigation. It serves as the QUIC baseline, isolating the effect of QUIC's transport mechanisms (encryption, congestion control) from stream multiplexing.

\textbf{Per-topic} (Fig.~\ref{fig:strategies}b) assigns each MQTT topic to a dedicated QUIC stream. Control packets (CONNECT, SUBSCRIBE, PINGREQ, etc.) use stream~0, while PUBLISH packets are routed to topic-specific streams. A hash map maintains the topic-to-stream mapping, creating streams lazily on first publish. This provides topic-level HOL blocking isolation: a retransmission delay on one topic's stream does not block delivery on other topics' streams.

\textbf{Per-publish} (Fig.~\ref{fig:strategies}c) creates a new unidirectional QUIC stream for each PUBLISH packet. This provides the finest-grained isolation---every message is independent---but at the highest overhead cost, as stream creation requires allocating QUIC state for each message.

\subsection{Flow Header Protocol}

When using per-topic or per-publish strategies, the receiver must determine how to demultiplex incoming data on non-control streams. We introduce a lightweight \emph{flow header} prepended to the MQTT payload on data streams.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\footnotesize]
\draw (0,0) rectangle (7, 0.8);

\draw (1.2,0) -- (1.2,0.8);
\draw (2.4,0) -- (2.4,0.8);
\draw (7.0,0) -- (7.0,0.8);

\node at (0.6, 0.4) {Type};
\node at (1.8, 0.4) {Flags};
\node at (4.7, 0.4) {MQTT Packet Data};

\node[font=\tiny, text=gray] at (0.6, -0.25) {1 byte};
\node[font=\tiny, text=gray] at (1.8, -0.25) {1 byte};
\node[font=\tiny, text=gray] at (4.7, -0.25) {variable};

\draw[decorate, decoration={brace, amplitude=4pt, mirror}] (0, -0.6) -- (7, -0.6)
  node[midway, below=5pt, font=\tiny] {QUIC stream payload};
\end{tikzpicture}
\caption{Flow header format. The 2-byte header identifies the packet type and flags, enabling the receiver to parse data streams without full MQTT framing on stream~0.}
\label{fig:flow-header}
\end{figure}

The flow header (Fig.~\ref{fig:flow-header}) consists of two bytes: a type field identifying the MQTT packet type (PUBLISH for data streams) and a flags byte encoding QoS level, retain, and DUP. The receiver reads this header to determine how to parse the remainder of the stream. This is more efficient than transmitting the full MQTT fixed header on every stream, since the stream itself provides the framing boundary.

\subsection{Datagram Transport}

As an alternative to stream-based transport, \texttt{mqtt5} supports QUIC datagrams (RFC~9221) for QoS~0 messages. Datagrams bypass QUIC's stream layer entirely: messages are sent as unreliable frames within the QUIC connection, benefiting from QUIC's encryption and connection management but without reliability or ordering guarantees.

Datagram transport is appropriate for latency-sensitive telemetry where occasional message loss is acceptable. Since QUIC datagrams are subject to the connection's congestion window but not to per-stream flow control, they achieve lower latency under loss while maintaining comparable throughput to stream transport (see Section~\ref{sec:evaluation}).

\subsection{Broker Integration}

The client and broker configure their stream strategies independently. On the client side, the application selects a strategy (control-only, per-topic, or per-publish) before connecting; this governs how outbound PUBLISH packets are dispatched onto QUIC streams. On the broker side, a server-wide delivery strategy governs how the broker sends messages to subscribers. In a typical deployment, both sides are configured to the same strategy, but the architecture does not enforce this---a per-topic client can connect to a control-only broker, or vice versa.

For outbound delivery, the broker's stream manager maintains a per-connection topic-to-stream mapping. When configured for per-topic delivery, streams are created lazily on first use and reused for subsequent messages on the same topic, with LRU eviction to bound stream count. For per-publish delivery, the broker opens a new unidirectional stream for each message. On the receive side, the broker's QUIC acceptor classifies incoming streams by ID: stream~0 carries the full MQTT packet parser, while non-zero streams use the flow header parser to extract PUBLISH packets and route them into the topic distribution engine.
