\begin{abstract}
MQTT, the dominant IoT messaging protocol, operates over TCP, where head-of-line (HOL) blocking couples logically independent topic flows: a single packet loss delays delivery across all topics sharing a connection. We define three configurable stream mapping strategies---control-only, per-topic, and per-publish---for running MQTT over QUIC, each offering a different point in the isolation--overhead tradeoff space. We implement these strategies in \texttt{mqtt5}, an open-source MQTT library, and evaluate across six experiments on GCP infrastructure under controlled network impairment. Using inter-topic latency spread to quantify HOL blocking, we find that per-topic QUIC amplifies spread by 192$\times$ under 5\% packet loss (from 0.09\,ms at baseline to 17.8\,ms), confirming independent per-stream loss recovery, while TCP spread increases only 1.3$\times$ (from 6.6\,ms to 8.5\,ms). Spike isolation ratio corroborates this: at 1\% loss, TCP exhibits 0.99 (near-perfect co-occurrence of latency spikes across topics) versus 0.72 for per-topic QUIC and 0.60 for per-publish. At baseline (0\% loss), TCP achieves 1.5--2.4$\times$ the throughput of QUIC strategies depending on strategy and connection count, but under loss ($\geq$1\%) QUIC outperforms TCP by up to 2.7$\times$ due to more efficient loss recovery. Per-topic mapping maintains throughput within 13--32\% of single-stream QUIC. We additionally evaluate QUIC datagram transport, finding a 19\% latency advantage over streams at 5\% packet loss with comparable throughput. These results provide concrete strategy selection guidelines for deploying MQTT over QUIC in IoT environments.
\end{abstract}
